---
title: "Exploring the BRFSS data"
author: "Raghavesh Viswanath"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
    df_print: paged
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(ggpubr)
library(gridExtra)
library(scales)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data

#### Data Collection Process

Spanning 50 U.S. states, the District of Columbia, and three U.S. territories, more than 400,000 adult BRFSS surveys are conducted each year. A randomly sampled adult from each participating household fills out a survey regarding health-related risk behaviors and events, chronic health conditions, and use of preventive services in order to gauge emerging health issues. Surveys are conducted over the phone, which leads to my concern that those in poverty unable to afford a landline/cell phone are unable to participate in the survey and moreover are potentially more at-risk due to not being able to afford health services, which could potentially skew the results of the questionnaire.

#### Generalizable Results

The data was gathered randomly sampling an adult from each household or residence. The study is observational because the researchers merely collected data without interfering; the researchers asked a questionnaire instead of overseeing an experiment. Therefore, we can infer from the observational nature of the study and the fact that participants were randomly sampled that the study's results are generalizable. 

#### Non-causal Results

No random assignment was used when gathering data as participants were never assigned into groups after being sampled. Therefore, we can further infer that the results of the study are non-causal and only correlation statements can be made using the results.

* * *

## Part 2: Research questions

**Research question 1:**

What is relation between income and education level of our  respondents?

Variables Required: income2, educa

**Research question 2:**

One of the first research question involves how sleeptime of an US citizen have any affects on chronic health condition specifically depressive disorder, specifically for those category of people who sleep for less than 7 hours.

Variables Required: sleptim1, addepev2

**Research question 3:**

Is there a correlation between being a veteran and poor physical or mental health, and can an association between both aforementioned variables and binge drinking be inferred?

Variables Required: veteran3,poorhlth 

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research question 1:**

Is there a time of year where the interview is not coded as fully completed? Would months with more holidays result in a change of completed interviews? Looking further into this, is there a state that sticks out for these cases?

Variables Required: income2, educa

#### Select Variables of Interest

Filtering out the data to work with our variables of interest
```{r}
q1<-data.frame(Income=brfss2013$income2, EducationLevel=brfss2013$educa)
T1<-table(q1)
T1<-data.frame(T1)
```

#### Variable Exploration
Getiing rid of all rows containing NA values as they contribute nothing to making inferences.
```{r}
T1<-na.omit(T1)
unique(T1$EducationLevel)
```
Seperating the data in 6 seperate data frames, where each one corresponds to respective Education Level and adding a Proportion Variable which will later be used to represent the frequency distribution using a pie-chart.
```{r}
H1<-subset(T1,T1$EducationLevel="Never attended school or only kindergarten")
H1<-cbind(H1,Prop=H1$Freq/colSums(H1[,3,drop=FALSE]))

H2<-subset(T1,T1$EducationLevel="Grades 1 through 8 (Elementary)")
H2<-cbind(H2,Prop=H2$Freq/colSums(H2[,3,drop=FALSE]))

H3<-subset(T1,T1$EducationLevel="Grades 9 though 11 (Some high school)")
H3<-cbind(H3,Prop=H3$Freq/colSums(H3[,3,drop=FALSE]))

H4<-subset(T1,T1$EducationLevel="Grade 12 or GED (High school graduate)")
H4<-cbind(H4,Prop=H4$Freq/colSums(H4[,3,drop=FALSE]))

H5<-subset(T1,T1$EducationLevel="College 1 year to 3 years (Some college or technical school)")
H5<-cbind(H5,Prop=H5$Freq/colSums(H5[,3,drop=FALSE]))

H6<-subset(T1,T1$EducationLevel="College 4 years or more (College graduate)")
H6<-cbind(H6,Prop=H6$Freq/colSums(H6[,3,drop=FALSE]))
```


We are expecting 53 values, however there are 55 in the table. Scrolling through the table we find 2 unexpected values; 0 and 80. 0 and 80 are not in the [State FIPS list](https://www.mcc.co.mercer.pa.us/dps/state_fips_code_listing.htm). Thus, these will be removed.



#### Exploration

I've chosen the viridis Color palette to represent the distribution in the pie-chart and allotocating grid space for the 6 pie-chart to be plotted.

```{r, message=F}
windows()
par(mfrow=c(2,3))
library("viridis")
```
Now ,plotting the corresponding pie-chart to make inferences
```{r}
g1<-ggplot(H1,aes(x="",y=Prop,fill=Income))+scale_color_viridis(option = "D") + ggtitle("Never attended school or only kindergarten")
g1+geom_bar(stat="identity",color="white")+coord_polar("y",start=0)+theme_void()+scale_color_viridis(discrete = TRUE, option = "D")+scale_fill_viridis(discrete = TRUE) 


g2<-ggplot(H2,aes(x="",y=Prop,fill=Income))+scale_color_viridis(option = "D") + ggtitle("Never attended school or only kindergarten")
g2+geom_bar(stat="identity",color="white")+coord_polar("y",start=0)+theme_void()+scale_color_viridis(discrete = TRUE, option = "D")+scale_fill_viridis(discrete = TRUE) 


g3<-ggplot(H3,aes(x="",y=Prop,fill=Income))+scale_color_viridis(option = "D") + ggtitle("Never attended school or only kindergarten")
g3+geom_bar(stat="identity",color="white")+coord_polar("y",start=0)+theme_void()+scale_color_viridis(discrete = TRUE, option = "D")+scale_fill_viridis(discrete = TRUE) 


g4<-ggplot(H4,aes(x="",y=Prop,fill=Income))+scale_color_viridis(option = "D") + ggtitle("Never attended school or only kindergarten")
g4+geom_bar(stat="identity",color="white")+coord_polar("y",start=0)+theme_void()+scale_color_viridis(discrete = TRUE, option = "D")+scale_fill_viridis(discrete = TRUE) 


g5<-ggplot(H5,aes(x="",y=Prop,fill=Income))+scale_color_viridis(option = "D") + ggtitle("Never attended school or only kindergarten")
g5+geom_bar(stat="identity",color="white")+coord_polar("y",start=0)+theme_void()+scale_color_viridis(discrete = TRUE, option = "D")+scale_fill_viridis(discrete = TRUE) 


g6<-ggplot(H6,aes(x="",y=Prop,fill=Income))+scale_color_viridis(option = "D") + ggtitle("Never attended school or only kindergarten")
g6+geom_bar(stat="identity",color="white")+coord_polar("y",start=0)+theme_void()+scale_color_viridis(discrete = TRUE, option = "D")+scale_fill_viridis(discrete = TRUE) 
```

We see from the pie-charts that as education level increases, you have more of chance of getting paid the big bucks. This is'nt something unexpected though, it just reiterates the fact that Higher Education is more likely to  provide you with High income than other means.
---

**Research question 2:**

One of the first research question involves how sleeptime of an US citizen have any affects on chronic health condition specifically depressive disorder, specifically for those category of people who sleep for less than 7 hours.

Variables Required: sleptim1, addepev2

#### Select Variables of Interest

```{r}
SLD <- dataframe(brfss2013$sleptim1,brfss2013$addepev2)
```

#### Variable Exploration

```{r}
SLD_clean<-na.omit(SLD)
up_dep<- SLD_clean %>%
  filter(sleptim1<5) %>%
  group_by(addepev2) %>%
  summarise(count=n())
```
Filtering the data to get what we need using the piping operations
#### Exploration

Plotting the data to make inferences.

```{r}
ggplot(data=up_dep,aes(x=addepev2,y=count))+geom_bar(stat="identity",color='blue',fill='green')+xlab("Depressive Order for people having <7 hours of sleep?")+ylab("Number of Respondents")
```
As normal Human Beings require atleast 7 hours of sleep daily, i wanted to know if less sleep than usual would lead to them having depressive disorder(pretty far fetched thought) Results indicate that majority dont report having a depressive disorder.

**Research question 3:**
Is there a correlation between being a veteran and poor physical or mental health?

Variables Required: veteran3, poorhlth

#### Select Variables of Interest
Filtering out rows containing NA as they are useless.
```{r}
q2tbl <- brfss2013 %>% filter(!(is.na(veteran3)), !(is.na(poorhlth)))
```

#### Variable Exploration

```{r}
q2tbl <- q2tbl %>% mutate(poorhlth = ifelse(poorhlth > 7, "8+", poorhlth))
q2tbl$poorhlth <- factor(q2tbl$poorhlth)
```
In the statistics above, we generalize days binge drank in the last month exceeding a week as '8+' to indicate excessive binge drinking. The summary statistics illustrate a choppy normal distribution of proportions of veterans binge drinking, with the distribution fluctuating up and down independent from the number of days binge drank, hinting at a possible correlation.

```{r}
q2tbl %>% group_by(poorhlth, veteran3) %>% summarize(count=n())
```



#### Exploration


```{r}
library("viridis")
q2plt <- ggplot(q2tbl) + aes(x = veteran3, fill = poorhlth) + geom_bar(position = "fill")+scale_color_viridis(option = "D")
q2plt <- q2plt +  + xlab("Veteran, ) + ylab("Relative Proportion")+  scale_color_viridis(discrete = TRUE, option = "D")+ scale_fill_viridis(discrete = TRUE)
q2plt

```
The plot illustrates a normal distribution for no days of poor health across 0 through 8+ days spent binge drinking in the last month and a an increasing linear trend for poor health across the same range Moreover, there were proportionately less veterans than non-veterans who reported little to no days of poor health consistently across 0 through 8+ days spent binge drinking and proportionately more veterans who reported more days of poor health than non-veterans consistent across the same range. This trend indicates that there is a strong correlation between being a veteran, poor physical or mental health, and binge drinking.

